NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-31 Manuscript/manuscriptPS.tex
Updates in text and figures
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21 and some other updates

    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\

@@

    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & ? & ? & \cite{??} \todoi{To be added by Ollila}  \\

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4} \\
  \todo{Lipid17 and MacRog results should be added into this plot.}
\end{figure}

The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
as previously observed also for PC lipids \cite{botan15}. The models for PS lipids perform generally less well than for
PC lipids in the previous study
(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
Therefore, interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

\subsection{Counterion binding and interactions between PC and PS headgroups}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
is reproduced only in Gromos-CKP simulations and in the MacRog simulation
with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The counterion binding and concomintant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
because the MacRog simulation with strongest sodium binding
(the lowest concentrations in bulk water) gives the same area per molecule
as CHARMM36ua simulation with significantly weaker counterion binding
affinity.
%CHARMM36, CHARMM36ua and Slipid models give
%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increase the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
%with potassium counterions and weaker
%binding affinity give larger area per molecule 
%increasing counterion binding affinity reduces the area per molecule
%when comparing between sodium and potassium in 
%or between
%ff99 and jc ions in Lipid 17 simulations.
In conclusion, the results are in line with the previous study suggesting that the
low area per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.
%, Berger and Lipid17/JC simulations.
%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
%counterion density, but slightly at different distances from the bilayer
%center. Other models exhibit more complicated distribution, often including
%two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Binding of coions to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of ion concentration \cite{catte16}. This is less straighforward
for charged lipid bilayers because counterions are always present and the
ion free reference state does not exist
(section \ref{mixtureTOadditionalCIs} in the supplementary information).
Here we evaluate the amount of bound charge using the changes of
headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
According to the electrometer concept, the headgroup order parameters of POPC
increase when negativaly charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions having
the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}),
while other simulations predict no change or decrease in the order parameters
(Fig. \ref{HGorderparametersPCvsPS}).
%In experiments and MacRog simulations, the increasing amount of negative charge
%in the membrane tilts the PC headgroup more parallel to the membrane plane
%leading to the larger order parameters.
In Berger and CHARMM36 simulations, the stronger counterion binding cancel
the effect of negatively charged headgroups and the headgroup order parameters
do not increase with increasing amount of PS lipids.
%the order parameters decrease with the increasing amount of
%positive counterions of PS lipid.
%In CHARMM36 simulations, the counterion binding neutralizes the negative charge
%of lipids at the interface and 
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
realistic surface charge density in membranes containing PS lipids,
while the other tested simulation models overestimate the counterion
binding affinity.
%reproduces the correct
%response of POPC headgroup to the increasing amount of negatively charged lipids,
%indicating that these simulations have 
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

The headgroup order parameteres of POPS shift closer to zero when bilayer
is diluted with POPC (Fig.~\ref{HGorderparametersPCvsPS}),
which is intereted to indicate less rigid structure of PS headgroups in
the mixture~\cite{browning80,buldt81,roux90,roux91,scherer87}.
This shift is observed only in lipid14/17 simulations
%, the headgroup order parameters of POPS shift closer
%to zero when the bilayer is diluted with POPC,
but the numerical values of order parameters are too far from experiments, having also different signs,
for the proper interpretation of the experimental data.
In CHARM36 and Gromos-CKP simulations, the shift of headgroup order
parameters toward zero are not observed when bilayer
diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
Therefore, we conclude that more accurate force fields are necessary
for MD simulation studies of PC-PS headgroup interactions.  
%The $\beta$-carbon order parameter
%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
%to the experimental data.
%The smaller $\alpha$-carbon order parameter increase in both
%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
%with increasing amount of PS, both model exhibiting a poor agreement with experiments.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
  \todo{Should we include also counterions into the plot?} \\
\end{figure}

>>>>

    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\

@@

    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & 80 & 120 & \cite{bergerPOPSPOPC4:1mixtureT310K}  \\

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids in the previous study
(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
Therefore, interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

\subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
is reproduced only in Gromos-CKP simulations and in the MacRog simulation
with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The counterion binding and concomintant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
because the MacRog simulation with strongest sodium binding
(the lowest concentrations in bulk water) gives the same area per molecule
as CHARMM36ua simulation with significantly weaker counterion binding
affinity.
%CHARMM36, CHARMM36ua and Slipid models give
%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
%with potassium counterions and weaker
%binding affinity give larger area per molecule 
%increasing counterion binding affinity reduces the area per molecule
%when comparing between sodium and potassium in 
%or between
%ff99 and jc ions in Lipid 17 simulations.
In conclusion, the results are in line with the previous study suggesting that the
low area per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.
%, Berger and Lipid17/JC simulations.
%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
%counterion density, but slightly at different distances from the bilayer
%center. Other models exhibit more complicated distribution, often including
%two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Binding of coions to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of ion concentration \cite{catte16}. This is less straighforward
for charged lipid bilayers because counterions are always present and the
ion free reference state does not exist.
In addition, the analysis is complicated by the artificial aggregation of counterions
in solution observed in some simulations (section \ref{mixtureTOadditionalCIs} in the supplementary information).
Here, we evaluate the amount of bound charge using the changes of
headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
According to the electrometer concept, the headgroup order parameters of POPC
increase when negativaly charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions having
the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}),
while other simulations predict no change or decrease in the order parameters
(Fig. \ref{HGorderparametersPCvsPS}).
%In experiments and MacRog simulations, the increasing amount of negative charge
%in the membrane tilts the PC headgroup more parallel to the membrane plane
%leading to the larger order parameters.
In Berger and CHARMM36 simulations, the stronger counterion binding cancel
the effect of negatively charged headgroups and the headgroup order parameters
do not increase with increasing amount of PS lipids.
%the order parameters decrease with the increasing amount of
%positive counterions of PS lipid.
%In CHARMM36 simulations, the counterion binding neutralizes the negative charge
%of lipids at the interface and 
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) predicts the most
realistic surface charge density in membranes containing PS lipids,
while the other tested simulation models overestimate the counterion
binding affinity. The results are in line with the changes of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the supplementary information.
%reproduces the correct
%response of POPC headgroup to the increasing amount of negatively charged lipids,
%indicating that these simulations have 
\begin{figure*}[]
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

The headgroup order parameteres of POPS shift closer to zero when bilayer
is diluted with POPC (Fig.~\ref{HGorderparametersPCvsPS}),
which is interpreted to indicate less rigid structure of PS headgroups in
the mixture~\cite{browning80,buldt81,roux90,roux91,scherer87}.
This shift is observed only in lipid14/17 simulations
%, the headgroup order parameters of POPS shift closer
%to zero when the bilayer is diluted with POPC,
but the numerical values of order parameters are too far from experiments, having also different signs,
for the proper interpretation of the experimental data.
In CHARM36 and Gromos-CKP simulations, the shift of headgroup order
parameters toward zero are not observed when bilayer
diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
Therefore, we conclude that more accurate force fields are necessary
for MD simulation studies of PC-PS headgroup interactions.  
%The $\beta$-carbon order parameter
%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
%to the experimental data.
%The smaller $\alpha$-carbon order parameter increase in both
%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
%with increasing amount of PS, both model exhibiting a poor agreement with experiments.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Figure needs general improvement.}
\end{figure}
